---
title: "Is the CFA Zone Less Developed than the Sub-Saharan Non-CFA Zone?"
author: "Salfo Bikienga"
date: '2019-11-03'
header:
  caption: ''
  image: ''
slug: is-the-cfa-zone-less-developed-than-the-sub-saharan-non-cfa-zone
tags:
- Monetary Economics
- FCFA
- Africa
- France
- Common Currency
categories: 
- CFA
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 7, fig.width = 9,
                      warning = FALSE, message = FALSE)
library(bookdown)
# http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/
# https://www.datacamp.com/community/tutorials/pca-analysis-r
```
```{r}
library(tidyverse)
library(hrbrthemes)
library(viridis)
library(grid)
library(pBrackets)
# https://www.data-to-viz.com/graph/barplot.html
# https://www.pauloldham.net/graphing-patent-data-with-ggplot2-part2/
```

If you have been following African issues on social media, you have probably come across a surge in anti-France sentiment in the Francophone zone, sometimes echoed in the Anglophone zone too. At the center of this sentiment is the common currency (known as the FCFA, or CFA) that 14 countries share.  

For some time, the sub-Saharan Francophone youth have been dancing to the rhythms of the anti-CFA drums, hammered by some Francophone African economists. Interestingly, we now have economists leading the struggle to liberate the continent of the remnants of the exploitative colonial relationship that France maintains with its former colonies. To make their case, these economists argue that the CFA zone is the least economically developed part of Africa. One can easily understand why economists are taking the lead in this struggle. After all, economics is all about alleviating human suffering by providing policy recommendations to improve the material well-being of people. Where else is the opinion of an informed economist more needed than in the least developed part of the least developed continent in the world?  

Being an economist from one of the CFA zone countries (Burkina Faso), I have decided to join the fight of a generation by lending my analytics skills to the anti-CFA discussion. Hopefully, no argument can beat powerful facts. So, let’s get the facts on whether the CFA zone is the least developed part of Africa, and whether the CFA is the culprit.  

The Human Development Index (HDI) is the main index used to rank countries in terms of economic development. It is a composite index that averages countries’ performance indexes in income, education and health. To my disappointment, though the HDI does indicate that a disproportionate number of the CFA zone countries are near the bottom in terms of development, it points to a culprit other than the CFA. The CFA is not the problem!

**Human Development Index**

The plot of the composite index, HDI, shows that the CFA zone performs poorly compared to the non-CFA zone (see Fig.1)^[The data are from the United Nations Development Programme, Human Development Reports' [website](http://hdr.undp.org/en/data). The latest available data, 2017, is used for this analysis. The latest available mortality rate is for 2016]. In fact, five of the 10 lowest-performing countries in Africa are in the CFA zone, despite the region's small representation (14 out of 48 sub-Saharan countries. Somalia is removed from the datasaet because of missing values). Clearly, the CFA zone is the least developed part of Africa. *Or, is it?* Let’s look at the individual elements (income, education and health) of the HDI to yank out the factors driving the CFA zone’s low rankings.

```{r}
cfa_labels <- read_csv("data/cfa/cfa_labels.csv")
life_exp <- read_csv("data/cfa/life_expectancy_at_birth.csv", skip = 1) %>%
  select(c("Country", "2017"))
mean_educ <- read_csv("data/cfa/mean_year_schooling.csv", skip = 1) %>%
  select(c("Country", "2017"))
gni <- read_csv("data/cfa/gni.csv", skip = 1) %>%
  select(c("Country", "2017"))
hdi <- read_csv("data/cfa/hdi.csv", skip = 1) %>%
  select(c("Country", "2017"))
infant_mort <- read_csv("data/cfa/infant_mortality.csv", skip = 1) %>%
  select(c("Country", "2016"))
mort_under5 <- read_csv("data/cfa/mortality_of_under5.csv", skip = 1) %>%
  select(c("Country", "2016"))

hdi_df <- left_join(cfa_labels, life_exp, by = "Country") %>%
  rename(life_exp = `2017` ) %>%
  left_join(mean_educ, by = "Country") %>%
  rename(mean_educ = `2017` ) %>%
  left_join(gni, by = "Country") %>%
  rename(gni = `2017` ) %>%
  left_join(hdi, by = "Country") %>%
  rename(hdi = `2017` ) %>%
  left_join(infant_mort, by = "Country") %>%
  rename(infant_mortality = `2016` ) %>%
  left_join(mort_under5, by = "Country") %>%
  rename(mortality_under5 = `2016` ) %>%
  filter(Country != "Somalia")

hdi_africa <- hdi_df %>%
  filter(sub_sahara == "yes")
hdi_africa$sub_sahara <- as.factor(hdi_africa$sub_sahara)
hdi_africa$cfa_zone <- as.factor(hdi_africa$cfa_zone)
hdi_africa$life_exp <- as.numeric(hdi_africa$life_exp)
hdi_africa$mean_educ <- as.numeric(hdi_africa$mean_educ)
hdi_africa$gni <- as.numeric(hdi_africa$gni)
hdi_africa$hdi <- as.numeric(hdi_africa$hdi)
hdi_africa$infant_mortality <- as.numeric(hdi_africa$infant_mortality)
hdi_africa$mortality_under5 <- as.numeric(hdi_africa$mortality_under5)
```

```{r}
old_names <- c("Central African Republic", "Congo (Democratic Republic of the)",
  "Equatorial Guinea", "Sao Tome and Principe", "Tanzania (United Republic of)")
new_names <- c("Cent. Afri. Rep.", "Congo (DRC)",
  "Equat. Guinea", "Sao Tome", "Tanzania")
hdi_africa[hdi_africa$Country == "Central African Republic", "Country"] = "Cent. Afri. Rep."
hdi_africa[hdi_africa$Country == "Congo (Democratic Republic of the)", "Country"] = "Congo (DRC)"
hdi_africa[hdi_africa$Country == "Equatorial Guinea", "Country"] = "Equat. Guinea"
hdi_africa[hdi_africa$Country == "Eswatini (Kingdom of)", "Country"] = "Eswatini"
hdi_africa[hdi_africa$Country == "Sao Tome and Principe", "Country"] = "Sao Tome"
hdi_africa[hdi_africa$Country == "Tanzania (United Republic of)", "Country"] = "Tanzania"
hdi_africa$cfa_zone <- ifelse(hdi_africa$cfa_zone == "yes", "cfa_zone", "non_cfa")
```



```{r, fig.cap="Ranking of sub-Saharan countries by human development index"}
bp_hdi <- ggplot(data = hdi_africa, aes(x = reorder(Country, hdi))) +
  geom_col(aes(y = hdi, fill = cfa_zone)) + 
  xlab("Country") + ylab("Human Development Index (HDI)") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by HDI") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum() +
  annotate("segment", x = c(12, 12, 12), xend = c(3, 8, 12), 
           y = 0.65, yend = 0.45, colour = "blue", lwd = 1,
             arrow = arrow(angle = 30, length = unit(2, "mm"), type = "closed")) +
    annotate("text", x = 12, y = 0.72, label = "Concentration \n of CFA countries \n at the bottom")
print(bp_hdi)
```

**Income**

Fig. 2 ranks the sub-Saharan countries by per-capita national income. Income remains one of the most used indexes of economic performance. The figure indicates that only two CFA countries are among the bottom 10 economies in sub-Saharan Africa. Wait! What is going on? The HDI indicates that the CFA countries are the poorest, but the per-capita income measure indicates otherwise. The CFA does not look bad at all! Are we fighting the wrong fight?


```{r, fig.cap="Ranking of sub-Saharan countries by per-capita national income"}
bp_gni <- ggplot(data = hdi_africa, aes(x = reorder(Country, gni))) +
  geom_col(aes(y = gni, fill = cfa_zone)) + 
  xlab("Country") + ylab("Per Capita Gross National Income") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by Income") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum() +
  annotate("segment", x = c(12, 12), xend = c(1, 5), 
           y = 10000, yend = 1000, colour = "blue", lwd = 1,
             arrow = arrow(angle = 30, length = unit(2, "mm"), type = "closed")) +
    annotate("text", x = 12, y = 12000, label = "Only 2 CFA\n countries are\n among the bottom 10")
print(bp_gni)
```

**Education**

The education index is one of the three elements of the HDI. Fig. 3 ranks the countries by the average years of schooling in each country. The plot indicates that the CFA zone countries are uncontested at the bottom of this category. Four of the five least educated countries in Africa (actually, in the world) are in the CFA zone. Is the CFA the reason people don’t go to school in that zone?


```{r, fig.cap="Ranking of sub-Saharan countries by average years of schooling"}
bp_educ <- ggplot(data = hdi_africa, aes(x = reorder(Country, mean_educ))) +
  geom_col(aes(y = mean_educ, fill = cfa_zone)) + 
  xlab("Country") + ylab("Mean Years of Education") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by Education") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum() +
  annotate("segment", x = c(12, 12), xend = c(2.5, 8.5), 
           y = c(6.5, 6.5), yend = c(2.3, 3.1), colour = "blue", lwd = 1,
             arrow = arrow(angle = 30, length = unit(2, "mm"), type = "closed")) +
    annotate("text", x = 12, y = 7.5, label = "6 CFA countries\n are among the\n bottom 10")
print(bp_educ)
```


```{r, eval=FALSE}
bp_life <- ggplot(data = hdi_africa, aes(x = reorder(Country, life_exp))) +
  geom_col(aes(y = life_exp, fill = cfa_zone)) + 
  xlab("Country") + ylab("Life Expectancy at Birth") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by Health") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum()
print(bp_life)
```


```{r, eval=FALSE}
bp_inf_Mort <- ggplot(data = hdi_africa, aes(x = reorder(Country, infant_mortality))) +
  geom_col(aes(y = infant_mortality, fill = cfa_zone)) + 
  xlab("Country") + ylab("Infant Mortality") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by Infant Mortality") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum()
print(bp_inf_Mort)
```

**Health**

It turns out that the CFA zone countries have the shortest life expectancy at birth in Africa. Indeed, Fig. 4 ranks countries by the mortality rate of their under-fives (the life expectancy measure, the main measure of health, shows the same picture). Wow! The CFA must be killing the people, literally. This is unsettling! If the CFA zone is not financially the most destitute part of Africa (see Fig. 2), why aren’t they educating their people? Why aren’t they preventing infant mortality? Do not accuse the culture! We need more sophisticated explanations.

```{r, fig.cap="Ranking of sub-Saharan countries by mortality rate of childreen under the age 5"}
bp_under5 <- ggplot(data = hdi_africa, aes(x = reorder(Country, mortality_under5))) +
  geom_col(aes(y = mortality_under5, fill = cfa_zone)) + 
  xlab("Country") + ylab("Mortality of the Under 5") +
  coord_flip() +
    ggtitle("CFA vs. Non CFA Zone: Ranking by Infant Mortality") +
  scale_fill_viridis(discrete = TRUE, name="", 
                     direction = 1, option = "D") +
  theme_ipsum() +
  annotate("segment", x = c(12, 12), xend = c(1, 10), 
           y = c(75, 75), yend = c(25, 50), colour = "blue", lwd = 1,
             arrow = arrow(angle = 30, length = unit(2, "mm"), type = "closed")) +
    annotate("text", x = 12, y = 90, label = "No CFA country\n among the top 10 \n achievers")
print(bp_under5)
```

To summarize, the CFA zone countries are disproportionately less developed than the non-CFA zone in sub-Saharan Africa. Though not the least wealthy, it is the least educated and least healthy region in the world. Too many of their children die young, and if they survive past five years old, they are less likely to receive a formal education than children in the rest of the world. Clearly, there is a fight to be fought. However, the CFA, France, and neocolonialism do not appear to be an economic urgency. The fight to be fought in this generation is not a fight against a “supposed” neocolonialism. It is a jihad! It is a fight against ourselves. Pointing our fingers at France will not keep our children alive, nor will it get them the education they need to operate in this world of artificial intelligence.
